---
title: "Metric Analysis"
author: "Mendoza College of Business, ITAO 40530"
date: "Spring 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

In this session we work with personal metric data, the same kind of self-tracked
daily measurements you collected (or are collecting) for Assignment 1. The
analysis has two goals:

1. Construct **lagged summary features** from a daily metric so we can ask
   questions like "does sleep over the last 3 days predict productivity today?"
2. Use those features in a small predictive model and inspect which metrics
   actually carry signal using SHAP.

The same code runs against either the provided class dataset (a full year of
metrics) or your own personal metric template. A single toggle at the top of
the file controls which one is loaded.

---

# Set Up

## Packages

```{r}
# install.packages("Hmisc")
# install.packages("matrixStats")
# install.packages("data.table")
# install.packages("ggplot2")
# install.packages("readxl")
# install.packages("xgboost")
# install.packages("shapviz")

library(Hmisc)        # Lag function (kept available for reference)
library(matrixStats)  # rowMaxs / rowMins for fast row-wise summaries
library(data.table)   # data.table and the data.table::melt used for plotting
library(ggplot2)      # plotting
library(readxl)       # reading the personal metric xlsx template
library(xgboost)      # gradient-boosted tree model
library(shapviz)      # SHAP explanations (replaces the older sourced helper file)
```

## Plot theme

Defining the palette and theme function once at the top means every plot in
this file shares the same look. The same `theme_hpa()` function is used in
later sessions (sourced from `explainer_functions.R`); we redefine it here
so this file can be run standalone.

```{r}
# ---- Dark mode color palette used throughout this session ----
dark_bg      <- "#0d1117"   # Deep background
dark_panel   <- "#161b22"   # Panel background (slightly lighter)
dark_grid    <- "#21262d"   # Subtle grid lines
dark_text    <- "#e6edf3"   # Primary text
dark_sub     <- "#8b949e"   # Subtitle / secondary text
dark_axis    <- "#30363d"   # Axis lines
accent_blue  <- "#58a6ff"   # Primary accent
accent_red   <- "#f85149"   # Danger / injury
accent_green <- "#3fb950"   # Success / healthy
accent_gold  <- "#d29922"   # Warning / moderate

# ---- Reusable dark theme for all plots in this session ----
theme_hpa <- function() {
  theme(
    plot.background = element_rect(fill = dark_bg, color = NA),
    panel.background = element_rect(fill = dark_panel, color = NA),
    panel.grid.major = element_line(color = dark_grid, linewidth = 0.3),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(color = dark_axis),
    axis.text = element_text(color = dark_sub, size = 9),
    axis.title = element_text(color = dark_text, size = 10),
    plot.title = element_text(face = "bold", color = dark_text, size = 13),
    plot.subtitle = element_text(color = dark_sub, size = 10),
    legend.background = element_rect(fill = dark_bg, color = NA),
    legend.key = element_rect(fill = dark_panel, color = NA),
    legend.text = element_text(color = dark_text, size = 9),
    legend.title = element_text(color = dark_text, size = 10),
    strip.background = element_rect(fill = dark_grid, color = NA),
    strip.text = element_text(color = dark_text, size = 10)
  )
}
```


## Loading data: class dataset or your own

The single switch below controls which dataset gets loaded. Leave it at `FALSE`
to follow along with the class data; flip it to `TRUE` once you have at least
a couple of weeks of entries in your personal metric template and want to run
the same pipeline on your own metrics.

```{r}
# ---- Data source toggle ----
# FALSE -> load the class metric dataset (metric_data_2020.rda)
# TRUE  -> load your filled-in Personal_Metric_Template.xlsx
USE_STUDENT_DATA <- TRUE

# ---- Paths (edit if your files live elsewhere) ----
class_data_path   <- "metric_data_2020.rda"
student_xlsx_path <- "04.15_metrics.xlsx"

# ---- Optional: rename the two custom columns from the template ----
# The template ships with generic "Metric 1" and "Metric 2" columns. If you
# tracked, e.g., caffeine and steps, set them here so the plots and SHAP
# output use the real names. Set to NULL to keep the defaults.
student_custom_names <- c("Metric 1" = "Caffeine_mg",
                          "Metric 2" = "Steps")
# student_custom_names <- NULL
```

The loader below normalizes whichever source you chose into a single data
frame called `metric_data`. Three things happen regardless of source:

  * Column names get spaces replaced with underscores (so `Wake Up` becomes
    `Wake_Up` and the rest of the file can reference columns by a single name).
  * The `Date` column is converted to a numeric `day_index`, days since the
    first observation. We use this index instead of row position when computing
    lagged values, which means gaps in your data (skipped days) are handled
    correctly: "3 days ago" really means 3 calendar days, not 3 rows back.
  * Fully blank rows (common in a freshly-printed template) are dropped.

```{r}
# ---- Helper: normalize a raw data frame into the shape the rest of the file expects ----
# This function takes whatever shape the source data arrives in (rda or xlsx)
# and returns a data frame with three guarantees: column names contain no
# spaces, every row has a numeric day_index, and there are no entirely-blank
# rows. By isolating this prep work in a function, the rest of the file can
# assume a clean input regardless of which source produced it.
prepare_metric_data <- function(raw, rename_map = NULL) {

  # ---- Step 1: Standardize column names ----
  # Spaces in column names force students to write metric_data[["Wake Up"]]
  # instead of metric_data$Wake_Up. Replacing spaces with underscores once
  # here lets us use the cleaner $ syntax everywhere downstream.
  names(raw) <- gsub(" ", "_", names(raw))

  # ---- Step 2: Apply optional renaming for student custom columns ----
  # The xlsx template has placeholder "Metric 1" / "Metric 2" columns. If the
  # student supplied real names via rename_map, swap them in. We loop because
  # there may be one, two, or zero renames depending on what the student set.
  if (!is.null(rename_map)) {
    map_from <- gsub(" ", "_", names(rename_map))
    map_to   <- unname(rename_map)
    for (k in seq_along(map_from)) {
      hit <- which(names(raw) == map_from[k])
      # which() returns integer(0) if no match, so length(hit) == 1 guards
      # against both "no column found" and "duplicate column names".
      if (length(hit) == 1) names(raw)[hit] <- map_to[k]
    }
  }

  # ---- Step 3: Build the day_index from the Date column ----
  # The lag calculations downstream need to know how many calendar days
  # separate two rows. We compute that here as an integer offset from the
  # earliest observation. If there's no Date column at all (unusual), we
  # fall back to row position so the rest of the pipeline still runs.
  if ("Date" %in% names(raw)) {
    # Drop rows where Date is missing, these are blank template rows
    raw <- raw[!is.na(raw$Date), , drop = FALSE]
    # Coerce to Date class regardless of how it was read in (POSIXct from
    # readxl, character from CSV, or already-Date from an rda)
    raw$Date <- as.Date(raw$Date)
    # Sort by date so that lag lookups based on day_index work correctly.
    # The match() lookups don't strictly require sorting, but sorting makes
    # head() / tail() inspections meaningful.
    raw <- raw[order(raw$Date), , drop = FALSE]
    # day_index = days since the earliest observation
    raw$day_index <- as.integer(raw$Date - min(raw$Date, na.rm = TRUE))
  } else {
    # Fall back to row position when no Date column exists
    raw$day_index <- seq_len(nrow(raw)) - 1L
  }

  # ---- Step 4: Drop fully blank rows ----
  # A freshly printed template has dozens of empty rows below the last entry.
  # If every metric column is NA in a row, that row contributes nothing.
  metric_cols <- setdiff(names(raw), c("Date", "Day", "day_index"))
  if (length(metric_cols) > 0) {
    # rowSums on a logical matrix counts non-NA cells; == 0 means every cell
    # in that row was NA.
    all_na <- rowSums(!is.na(raw[, metric_cols, drop = FALSE])) == 0
    raw <- raw[!all_na, , drop = FALSE]
  }

  # Reset row names so subsequent indexing is predictable. Without this we'd
  # carry around the row numbers from the original sheet, which can confuse
  # printouts and joins.
  rownames(raw) <- NULL
  raw
}

# ---- Actually load the data ----
# The if/else picks the right loader based on the toggle. Note that load()
# silently brings the object metric_data into the environment by name, which
# is why we don't assign its return value to anything. read_xlsx() is the
# modern equivalent that returns a value directly.
if (USE_STUDENT_DATA) {
  #raw <- read_excel("x04.15_metrics.xlsx")
  raw <-x04_15_metrics 
  metric_data <- prepare_metric_data(as.data.frame(raw),
                                     rename_map = student_custom_names)
  cat("Loaded student data:", nrow(metric_data), "rows,",
      ncol(metric_data) - 3, "metrics.\n")
} else {
  load(class_data_path)  # introduces an object called metric_data
  metric_data <- prepare_metric_data(metric_data)
  cat("Loaded class data:", nrow(metric_data), "rows,",
      ncol(metric_data) - 3, "metrics.\n")
}

tail(metric_data)
```

## Detecting which metrics are present

Rather than hardcoding the list of metrics, we figure it out from the loaded
data. Anything that isn't `Date`, `Day`, or `day_index`, and is numeric,
counts as a metric we can analyze. This pattern (detect columns from the data
rather than hardcoding) is what lets the same downstream code handle both the
9-column class dataset and a 5-column student tracker without modification.

```{r}
# ---- Identify metric columns automatically ----
# Step 1: list every column except the bookkeeping columns
non_metric_cols <- c("Date", "Day", "day_index")
metric_cols <- setdiff(names(metric_data), non_metric_cols)

# Step 2: keep only the numeric ones. sapply returns a logical vector here
# because is.numeric returns a single TRUE/FALSE per column. Filtering on
# this guards against, e.g., a stray text column the student added.
metric_cols <- metric_cols[sapply(metric_data[, metric_cols, drop = FALSE], is.numeric)]

cat("Detected metrics:\n  ", paste(metric_cols, collapse = ", "), "\n")
```

---

# Exploring the Raw Data

Before we start building features, it's worth looking at what we actually
have. Two questions matter most for a self-tracked dataset:

  1. **Where are the gaps?** Personal tracking is never complete. Knowing
     which metrics are missing when will shape how we interpret the model.
  2. **What do these metrics look like over time?** Daily values are noisy.
     Seeing the underlying trend (via a rolling average) is what motivates
     building lagged summary features in the next section.

## Data completeness heatmap

Each row below is a metric, each column is a date, and the cell color tells
you whether that value was recorded. This plot is the single most useful
diagnostic for a personal-tracking dataset because it exposes structural
gaps (metrics started late, metrics that drop off on weekends, travel
periods where nothing was logged) that would otherwise silently corrupt
downstream analysis.

```{r fig.height=4, fig.width=10}
# ---- Build a long-form presence indicator ----
# For the heatmap we need one row per (Date, metric) combination with a
# flag for whether the value is present. melt() handles the reshape; we
# use data.table because it's already loaded and it's fast.
completeness_dt <- as.data.table(metric_data[, c("Date", metric_cols)])
completeness_long <- melt(completeness_dt, id.vars = "Date",
                          variable.name = "metric", value.name = "value")

# is_present = TRUE when the value is not NA. This becomes the fill color.
completeness_long[, is_present := !is.na(value)]

# ---- Plot ----
# geom_tile gives us the calendar grid. We reverse the y axis so metrics
# appear in the order they're listed in metric_cols (tile plots otherwise
# show the first level at the bottom).
ggplot(completeness_long, aes(x = Date, y = metric, fill = is_present)) +
  geom_tile(color = dark_bg, linewidth = 0.05) +
  scale_fill_manual(values = c(`TRUE` = accent_green, `FALSE` = dark_grid),
                    labels = c(`TRUE` = "Recorded", `FALSE` = "Missing"),
                    name = NULL) +
  scale_y_discrete(limits = rev(metric_cols)) +
  labs(x = NULL, y = NULL,
       title = "Data completeness by metric and date",
       subtitle = "Green cells are recorded values, dark cells are missing") +
  theme_hpa() +
  theme(legend.position = "top")
```

Things to look for in your own version of this plot: metrics that only
started being recorded partway through, weekend gaps in evening metrics
(Productivity and Energy often drop off when work rhythm changes), and
long horizontal stretches of missing data that indicate periods to treat
with extra skepticism in the modeling section.

## Metric time series with 7-day rolling average

Raw daily values are noisy enough that trends are hard to see by eye. A
rolling mean smooths through the day-to-day variation and reveals the
underlying trajectory. This plot also previews what the lag-summary features
are doing mathematically: the `mean_val` column from our lag calculator is
effectively this rolling mean.

```{r fig.height=8, fig.width=10}
# ---- Compute a 7-day rolling mean for every metric ----
# We already have a function that computes this: lag_metric_calculator
# with lagged_days = 6 and include_current = TRUE gives us a 7-day window.
# Rather than writing a separate rolling-mean helper, we reuse the function
# we'll soon use for modeling. This is a small payoff for the effort of
# building the reusable function earlier.
#
# Note: this chunk runs before lag_metric_calculator is defined in the
# walkthrough below, so we inline a minimal version here. In practice you'd
# put the function definition before either use, but keeping it inline here
# lets the walkthrough stand on its own.
rolling_mean_7 <- function(vec, day_index) {
  # Build a matrix with 7 columns (today + 6 prior days)
  res_data <- matrix(NA_real_, nrow = length(vec), ncol = 7)
  res_data[, 7] <- vec  # today in the rightmost column
  for (i in 1:6) {
    target_days <- day_index - i
    match_idx   <- match(target_days, day_index)
    res_data[, i] <- vec[match_idx]
  }
  rowMeans(res_data, na.rm = TRUE)
}

# ---- Build long-form data with both raw values and rolling mean ----
ts_list <- list()
for (m in metric_cols) {
  vec <- metric_data[[m]]
  ts_list[[m]] <- data.frame(
    Date   = metric_data$Date,
    metric = m,
    raw    = vec,
    smooth = rolling_mean_7(vec, metric_data$day_index)
  )
}
ts_df <- do.call(rbind, ts_list)

# ---- Plot: raw values as faint points, rolling mean as gold line ----
# Faceting per metric with free_y scales because each metric has its own
# range (Sleep_Hours is 0-10, Stress is 1-5, etc.). ncol = 2 gives a
# readable layout for 9 metrics.
ggplot(ts_df, aes(x = Date)) +
  geom_point(aes(y = raw), color = accent_blue, alpha = 0.25, size = 0.8) +
  geom_line(aes(y = smooth), color = accent_gold, linewidth = 0.7) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  labs(x = NULL, y = "Metric value",
       title = "Daily metrics over time with 7-day rolling average",
       subtitle = "Blue points are daily values, gold line is the 7-day mean") +
  theme_hpa()
```

The gold line is doing the same work that the `mean_val` feature will do
in the model: averaging out day-to-day noise to expose the trend. If your
rolling mean line is nearly flat for a metric, that metric has little
medium-term variation to predict from; if it shows clear drift or cycles,
there's signal the model can potentially use.

---

# Lagged Feature Construction

A single bad night of sleep rarely shows up in next-day performance, but a
*string* of bad nights usually does. To capture that we need features that
summarize a metric over a recent window: its mean, sum, max, min, and how
many of those days are missing.

The lag function below uses the `day_index` column rather than row position.
This matters whenever the dataset has gaps: if you skipped logging on Tuesday,
"yesterday's value" on Wednesday should be Monday's, not blank.

## Walking through the calculation for one metric

We start by writing the calculation out for a single metric (`Sleep_Hours`),
one window (7 days), with the current day included. Once that works we wrap
it in a function. Doing it this way makes the loop logic visible before it
gets buried inside a function body.

```{r}
# ---- Parameters for the worked example ----
# These four values fully specify a single lag-summary calculation. They
# become the function arguments in the next chunk.
lagged_days <- 7              # how many calendar days back to include
metric <- "Sleep_Hours"       # which column to summarize
include_current <- TRUE       # include today's value as the rightmost column?

# ---- Pull out the metric vector and the day index it lines up with ----
# unlist() flattens the single-column data frame slice into a plain vector.
# Without it, met_vec would still be a one-column data frame and the rest
# of the math would silently produce something unexpected.
met_vec   <- unlist(metric_data[, metric])
day_index <- metric_data$day_index

# ---- Allocate the result matrix ----
# Each column holds the metric value at one lag offset. If we include today,
# we need lagged_days + 1 columns; otherwise just lagged_days. NA_real_ is
# used (rather than plain NA) so the matrix is numeric from the start; this
# avoids R having to coerce the type later when we assign numeric values.
n_cols <- if (include_current) lagged_days + 1 else lagged_days
res_data <- matrix(NA_real_, nrow = nrow(metric_data), ncol = n_cols)

# Today's value goes in the rightmost column when include_current = TRUE.
# This convention (today on the right, oldest on the left) is arbitrary but
# kept consistent across every call.
if (include_current) {
  res_data[, n_cols] <- met_vec
}

# ---- Fill in the lagged columns ----
# This is the heart of the day_index-based lag. For each lag i:
#   1. Compute the target day_index for every row (today minus i days).
#   2. Use match() to find the row position (if any) with that day_index.
#   3. Pull the metric value from that position.
# match() returns NA whenever no such row exists, which is exactly the
# behavior we want for the first few rows of the dataset and for any
# genuine gaps in tracking.
for (i in seq_len(lagged_days)) {
  target_days <- day_index - i
  match_idx   <- match(target_days, day_index)
  res_data[, i] <- met_vec[match_idx]
}

# ---- Compute row-wise summaries across the lag window ----
# rowMeans / rowSums are base R; rowMaxs / rowMins come from matrixStats
# and are much faster than apply(res_data, 1, max). na.rm = TRUE means a
# row with one missing day still produces a usable summary.
mean_val     <- rowMeans(res_data, na.rm = TRUE)
sum_val      <- rowSums(res_data, na.rm = TRUE)
max_val      <- rowMaxs(res_data, na.rm = TRUE)
min_val      <- rowMins(res_data, na.rm = TRUE)
missing_vals <- rowSums(is.na(res_data))

# ---- Repair edge case: rows that are entirely missing ----
# When every value in a row is NA, rowMaxs returns -Inf and rowMins returns
# +Inf (because the max of an empty set is, by convention, -Inf). Those
# infinite values would crash plotting code downstream, so we coerce them
# back to NA here.
max_val[is.infinite(max_val)] <- NA
min_val[is.infinite(min_val)] <- NA

results <- data.frame(mean_val, sum_val, max_val, min_val, missing_vals)
head(results)
tail(results)
```

## Wrapping it in a function

The same calculation will be needed for every metric and every lag window,
which is a lot of repetition. Wrapping it in a function makes the modeling
loop further down readable and removes the risk of typos creeping into one
copy of the code but not another.

```{r}
# ---- Reusable lag-window summary function ----
lag_metric_calculator <- function(metric_data,
                                  lagged_days = 3,
                                  metric = "Sleep_Hours",
                                  include_current = TRUE) {
  #'
  #' Compute summary statistics over a rolling window of past values for a
  #' single metric, using calendar-day spacing rather than row position.
  #'
  #' @param metric_data data frame with a numeric `day_index` column
  #' @param lagged_days how many calendar days back to include in the window
  #' @param metric column name of the metric to summarize
  #' @param include_current if TRUE, today's value is included in the window
  #'
  #' @return data frame with mean, sum, max, min, and count of missing values
  #'         across the window for every row in `metric_data`
  #'

  # Pull out the vectors we need (see the worked example above for why
  # unlist is required here).
  met_vec   <- unlist(metric_data[, metric])
  day_index <- metric_data$day_index

  # Allocate result matrix and fill in today's value if requested.
  n_cols <- if (include_current) lagged_days + 1 else lagged_days
  res_data <- matrix(NA_real_, nrow = nrow(metric_data), ncol = n_cols)
  if (include_current) res_data[, n_cols] <- met_vec

  # Fill the lagged columns by matching on day_index (gap-aware lag).
  for (i in seq_len(lagged_days)) {
    target_days <- day_index - i
    match_idx   <- match(target_days, day_index)
    res_data[, i] <- met_vec[match_idx]
  }

  # Row-wise summaries
  mean_val     <- rowMeans(res_data, na.rm = TRUE)
  sum_val      <- rowSums(res_data, na.rm = TRUE)
  max_val      <- rowMaxs(res_data, na.rm = TRUE)
  min_val      <- rowMins(res_data, na.rm = TRUE)
  missing_vals <- rowSums(is.na(res_data))

  # Replace infinite values from all-NA rows with NA
  max_val[is.infinite(max_val)] <- NA
  min_val[is.infinite(min_val)] <- NA

  data.frame(mean_val, sum_val, max_val, min_val, missing_vals)
}
```

## Computing summaries at three window lengths

We pick three windows that span very different timescales: the last day
(very short-term effects), a 3-day window (a long weekend's worth of
recovery), and a week. Comparing how the same metric behaves across these
three windows is what tells us which timescale carries the signal.

```{r}
# ---- Build sleep summaries at three window lengths ----
# Each call returns a 5-column data frame: mean, sum, max, min, missing_vals
# computed over a window of (lagged_days + include_current) days.
lag_2_res <- lag_metric_calculator(metric_data, lagged_days = 1,
                                   metric = "Sleep_Hours", include_current = TRUE)
lag_3_res <- lag_metric_calculator(metric_data, lagged_days = 2,
                                   metric = "Sleep_Hours", include_current = TRUE)
lag_7_res <- lag_metric_calculator(metric_data, lagged_days = 6,
                                   metric = "Sleep_Hours", include_current = TRUE)
```

---

# Visualizing Lagged Sleep vs. Other Metrics

We want to know whether recent sleep predicts how we feel or perform today.
A quick way in is to plot each summary statistic against another metric.
The helper below avoids repeating the same plotting code six times.

```{r}
# ---- Reusable plotting function ----
# Takes a lag-summary data frame, a response vector, and labels, and produces
# a faceted scatter plot with a smoother for each summary stat. Pulling this
# into a function means we can swap the response variable (Productivity, Wake
# Up, etc.) without copy-pasting twenty lines of ggplot code each time.
plot_lag_vs_metric <- function(lag_res, response_col, response_name, lag_label) {
  # ---- Combine the lagged summaries with the response column ----
  plot_df <- cbind(lag_res, response = response_col)

  # ---- Reshape to long form for faceting ----
  # We need one row per (observation, summary_stat) pair so ggplot can facet
  # by the summary stat. data.table::melt is fast and avoids a dependency on
  # the older reshape / reshape2 packages.
  plot_dt   <- as.data.table(plot_df)
  plot_long <- melt(plot_dt, id.vars = "response",
                    variable.name = "summary_stat", value.name = "value")

  # ---- Build the plot ----
  # Points are blue (calm, "data") and the smoother is gold (drawing the eye
  # to the trend). Faceting uses free scales because mean, sum, max, and min
  # live on very different ranges.
  ggplot(plot_long, aes(x = value, y = response)) +
    geom_point(alpha = 0.4, color = accent_blue) +
    geom_smooth(method = "loess", se = FALSE, color = accent_gold, linewidth = 0.8) +
    facet_wrap(~ summary_stat, scales = "free") +
    labs(x = "Sleep summary value",
         y = response_name,
         title = paste(response_name, "vs. lagged Sleep Hours,", lag_label)) +
    theme_hpa() +
    theme(legend.position = "none")
}
```

## Productivity vs. lagged sleep

The interesting question here is whether the *window length* changes the
picture. If only the last day matters, the 1-day plot will show a relationship
the others do not; if accumulated debt matters more, the 7-day window will
look cleanest.

```{r}
# Productivity is present in both the class dataset and the student template,
# but if a student deleted the column we don't want to crash. The %in% check
# is cheap insurance.
if ("Productivity" %in% names(metric_data)) {
  print(plot_lag_vs_metric(lag_2_res, metric_data$Productivity, "Productivity", "2-day window"))
  print(plot_lag_vs_metric(lag_3_res, metric_data$Productivity, "Productivity", "3-day window"))
  print(plot_lag_vs_metric(lag_7_res, metric_data$Productivity, "Productivity", "7-day window"))
} else {
  cat("No Productivity column in this dataset, skipping productivity plots.\n")
}
```

## Wake Up vs. lagged sleep

Wake Up is a self-rated quality of waking. We'd hope (but should not assume)
that more sleep leads to easier mornings. Worth noting: if `include_current
= TRUE` for the sleep window, this is partly tautological since last night's
sleep is what produced this morning's Wake Up score. The gap between the
1-day and 7-day windows is the more informative comparison.

```{r}
if ("Wake_Up" %in% names(metric_data)) {
  print(plot_lag_vs_metric(lag_2_res, metric_data$Wake_Up, "Wake Up", "2-day window"))
  print(plot_lag_vs_metric(lag_3_res, metric_data$Wake_Up, "Wake Up", "3-day window"))
  print(plot_lag_vs_metric(lag_7_res, metric_data$Wake_Up, "Wake Up", "7-day window"))
} else {
  cat("No Wake_Up column in this dataset, skipping wake-up plots.\n")
}
```

---

# Predictive Modeling

The plots give us intuition for individual metrics. To handle all of them at
once, and to let interactions between metrics surface, we move to a model.
Gradient-boosted trees handle small tabular datasets with mixed missingness
gracefully, which is why XGBoost is a reasonable default here.

## Building the full feature matrix

For every detected metric and every lag window, we generate the same five
summary stats. `Sleep_Hours` uses `include_current = TRUE` because we usually
know last night's sleep before today's productivity is recorded. Other
metrics use `include_current = FALSE` so we aren't predicting Productivity
from itself.

```{r}
# ---- Modeling parameters ----
# response_metric: the column we're trying to predict. Productivity is the
# default; change this to predict Mood, Energy, or one of your custom
# metrics. The rest of the modeling code reads from this variable, so you
# only need to change it in one place.
response_metric <- "Productivity"

# lags: the window lengths (in days back) we'll build features for. Using
# multiple windows lets the model figure out which timescale matters for
# each metric on its own.
lags <- c(1, 3, 7)

# feature_metrics: every numeric metric except the response. Including the
# response in the feature set would leak the answer (today's productivity
# value would be available as a feature for predicting today's productivity),
# so we explicitly remove it.
feature_metrics <- setdiff(metric_cols, response_metric)
```

```{r}
# ---- Construct the feature matrix ----
# For each (lag, metric) combination, compute the 5 summary stats and tag
# the column names with the metric name and lag length. The result is a
# wide matrix where each column is interpretable from its name alone:
#   mean_val_Sleep_Hours_7  = mean Sleep_Hours over the last 7 days
#   max_val_Stress_3        = max Stress over the last 3 days (excluding today)
res_list <- list()

for (i in seq_along(lags)) {
  for (j in seq_along(feature_metrics)) {
    m <- feature_metrics[j]
    # Sleep is the one metric where today's value (last night) is known
    # in advance of the response, so we include it in its own window.
    inc_current <- identical(m, "Sleep_Hours")
    temp <- lag_metric_calculator(metric_data,
                                  lagged_days = lags[i],
                                  metric = m,
                                  include_current = inc_current)
    # Tag the column names so the SHAP plot is self-documenting
    names(temp) <- paste(names(temp), m, lags[i], sep = "_")
    res_list[[length(res_list) + 1]] <- temp
  }
}

# do.call(cbind, res_list) is the base-R equivalent of list.cbind from rlist.
# It stitches every data frame in res_list side by side into one wide matrix.
m_dat <- do.call(cbind, res_list)
cat("Feature matrix:", nrow(m_dat), "rows by", ncol(m_dat), "columns.\n")
```

## Setting the response and dropping rows we can't model

```{r}
# Pull out the response vector. Using [[ ]] returns a vector directly,
# whereas $response_metric would look for a column literally called
# "response_metric" (which doesn't exist).
response <- metric_data[[response_metric]]

# Drop any row where the response is missing. We can't train on rows where
# we don't know the answer, and we can't validate against them either.
keep <- !is.na(response)
m_dat    <- m_dat[keep, , drop = FALSE]
response <- response[keep]

cat("After dropping missing-response rows:", nrow(m_dat), "rows remain.\n")
```

## Guarding against tiny datasets

If you're running this against your own data after a couple of weeks of
tracking, you won't have enough rows to train a deep boosted model. Rather
than letting XGBoost overfit silently, we scale `nrounds` down to whatever
the data can support and stop with a clear message if there isn't enough
to fit anything sensible.

```{r}
# ---- Adapt model size to dataset size ----
n_obs <- nrow(m_dat)

# Hard floor: below 15 usable rows there's no point trying. The error
# message tells the student exactly what to do (collect more data) rather
# than producing a meaningless model.
if (n_obs < 15) {
  stop("Only ", n_obs, " usable rows. Not enough to fit a model. ",
       "Add more entries to your tracker (aim for at least three weeks) and try again.")
}

# Heuristic: roughly 2 rounds per observation, capped at 100 and floored at
# 20. The cap prevents long training times on the full year of class data;
# the floor ensures even small datasets get enough rounds to find any
# signal at all.
n_rounds <- min(100, max(20, floor(n_obs * 2)))

# Tree depth: shallower trees on small data reduce overfitting. The
# threshold of 60 rows is a rough rule of thumb, not a hard cutoff.
max_depth <- if (n_obs < 60) 3 else 6

cat("Training XGBoost with nrounds =", n_rounds,
    "and max_depth =", max_depth, "on", n_obs, "rows.\n")
```

## Fitting the model

```{r}
# ---- Build the XGBoost data structure ----
# xgb.DMatrix is XGBoost's internal format. It bundles features and labels
# together and pre-processes them for fast access during training. The
# label argument is the response vector.
dtrain <- xgb.DMatrix(data = as.matrix(m_dat), label = response)

# ---- Train using the modern xgb.train API ----
# We use xgb.train (rather than the older xgboost() wrapper) because it
# accepts a structured params list, which is the convention in every other
# session of this course.
fit_1 <- xgb.train(
  params = list(
    objective = "reg:squarederror",  # standard squared-error regression
    max_depth = max_depth,            # set above based on dataset size
    eta = 0.1                         # learning rate; small + many rounds = stable
  ),
  data = dtrain,
  nrounds = n_rounds,
  verbose = 1,
  print_every_n = 20
)
```

## How well does the model actually fit?

Before we interpret SHAP values, we need to know whether the model is worth
interpreting. A SHAP plot will happily rank features by importance even for
a model that explains almost none of the variance, so we check overall fit
first. Two diagnostics are useful:

  * **Predicted vs. actual scatter.** Points should cluster around the 45°
    line if the model is tracking the target. Systematic departures (the
    model consistently under-predicting high values, for example) are
    visible here but hidden in aggregate metrics.
  * **RMSE and R².** RMSE is in the units of the response (productivity
    points) and answers "how far off is a typical prediction?" R² is the
    share of variance the model captures, roughly comparable across
    response variables.

```{r fig.height=5, fig.width=6}
# ---- Generate in-sample predictions ----
# These are predictions on the same data we trained on, so they'll look
# better than the model would look on held-out data. That's an honest
# caveat to share with students: a good in-sample fit is necessary but
# not sufficient for a useful model. For small personal datasets we
# usually don't have enough data to hold out a meaningful test set.
preds <- predict(fit_1, newdata = dtrain)

# ---- Compute summary fit statistics ----
# RMSE: square root of mean squared error, in response units
rmse <- sqrt(mean((response - preds)^2))
# R-squared: 1 - (residual variance / total variance)
r2   <- 1 - sum((response - preds)^2) / sum((response - mean(response))^2)

cat("In-sample RMSE:", round(rmse, 3),
    "| R-squared:", round(r2, 3), "\n")

# ---- Build the scatter plot data ----
fit_df <- data.frame(actual = response, predicted = preds)

# ---- Plot predicted vs actual with a 45-degree reference line ----
# The reference line is where points would land if every prediction were
# perfect. geom_abline adds it independent of the data range, so it
# extends across the whole plot area.
ggplot(fit_df, aes(x = actual, y = predicted)) +
  geom_abline(slope = 1, intercept = 0, color = dark_sub,
              linetype = "dashed", linewidth = 0.5) +
  geom_point(color = accent_blue, alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = accent_gold,
              linewidth = 0.7) +
  labs(x = paste("Actual", response_metric),
       y = paste("Predicted", response_metric),
       title = "Predicted vs. actual",
       subtitle = paste0("In-sample RMSE = ", round(rmse, 3),
                         ", R-squared = ", round(r2, 3))) +
  theme_hpa()
```

Interpreting the plot:

  * **Tight clustering along the dashed line** means the model captures
    most of the variation in the response.
  * **A shallow gold line relative to the dashed line** means the model
    is under-reactive: it predicts values closer to the mean than they
    actually are. This is normal for gradient-boosted models on small
    datasets because they regularize toward the global average.
  * **Points far from the line** are observations the model couldn't
    explain with the features we built. Looking at what days those are
    (holidays, illness, unusual events) often suggests features worth
    adding.

If R² is low (say, below 0.2), treat the SHAP plot that follows as
descriptive rather than prescriptive. Features may rank in a particular
order, but the model doesn't explain enough of the response for that
ranking to support strong claims.

## SHAP: which features actually carry signal?

`shapviz` computes SHAP values directly from the fitted model and the
feature matrix, then gives us beeswarm and importance plots out of the box.
Each dot in the beeswarm is one observation; horizontal position is that
feature's contribution to the model's prediction for that day. Color encodes
the feature's value (high = red, low = blue), so you can see at a glance
whether *high* values of a given lagged metric push predicted productivity
up or down.

```{r}
# ---- Construct the shapviz object ----
# shapviz takes the fitted model and the feature matrix used to train it
# and computes SHAP values for every observation. The X_pred matrix must
# be in the same column order and have the same column names as the
# training data.
sv <- shapviz(fit_1, X_pred = as.matrix(m_dat))

# ---- Beeswarm: distribution of SHAP values across observations ----
# Each row in the plot is one feature; each dot is one day. Dot position
# (left vs right) is the SHAP contribution; dot color is the feature value
# itself. The combination tells you both how important a feature is and
# the direction of its effect.
sv_importance(sv, kind = "beeswarm", max_display = 15) +
  scale_color_gradient(low = accent_blue, high = accent_red) +
  labs(title = paste("SHAP feature importance, predicting", response_metric)) +
  theme_hpa()

# ---- Bar chart: mean absolute SHAP value ----
# Easier to scan when you just want a ranking. The bar length is the average
# magnitude of each feature's contribution, regardless of direction.
sv_importance(sv, kind = "bar", max_display = 15, fill = accent_red) +
  labs(title = paste("Mean |SHAP|, predicting", response_metric)) +
  theme_hpa()
```

## SHAP dependence: how does the top feature affect predictions?

The beeswarm and bar charts tell us *which* features matter. The dependence
plot tells us *how* a feature matters. For a single feature, it plots the
feature's value on the x axis and its SHAP contribution on the y axis,
with one point per observation.

Reading a dependence plot answers questions the ranking can't:

  * **Is the relationship monotonic?** A clean upward slope means more is
    always better (or worse); a curve that flattens means there's a
    saturation point past which the feature stops mattering.
  * **Is there a threshold?** Sometimes a feature has no effect until it
    crosses a specific value, then a large effect above it. That shows up
    as a step function in the dependence plot.
  * **Is there interaction?** `sv_dependence` color-codes points by a
    second feature it thinks interacts strongly with the main one, so you
    can see whether the effect of sleep depends on, say, stress level.

```{r fig.height=4, fig.width=7}
# ---- Identify the top-ranked feature programmatically ----
# Rather than hardcoding "mean_val_Sleep_Hours_7" (which might not be the
# top feature on student data), we pull the name with the largest mean
# absolute SHAP value directly from the shapviz object. shapviz stores
# SHAP values in sv$S as a matrix with one column per feature.
mean_abs_shap <- colMeans(abs(sv$S))
top_feature   <- names(sort(mean_abs_shap, decreasing = TRUE))[1]
cat("Top feature by mean |SHAP|:", top_feature, "\n")

# ---- Dependence plot for the top feature ----
# By default sv_dependence picks a second feature to color by using
# SHAP interaction strength. The color scale ends up on whatever that
# feature's natural range is, so we let shapviz pick the palette rather
# than forcing our dark-theme colors.
sv_dependence(sv, v = top_feature) +
  labs(title = paste("SHAP dependence:", top_feature),
       subtitle = "X = feature value, Y = SHAP contribution, color = interacting feature") +
  theme_hpa()
```

A common pattern in this kind of data: the dependence plot for
`mean_val_Sleep_Hours_7` often shows a roughly S-shaped curve, flat below
some low threshold, rising through a middle zone, and flattening again
past the point of diminishing returns. That shape is exactly what a linear
model would miss and exactly what gradient-boosted trees can pick up.

## Reading the output

Two questions to ask of the SHAP plot:

1. **Which metrics dominate?** If recent sleep summaries (mean, sum) sit at
   the top, that's evidence the lagged-sleep story we built up visually
   actually holds when other metrics are competing for attention.
2. **Which window length wins?** Compare the same metric across lag tags
   (`_1`, `_3`, `_7`). The window the model leans on tells you the timescale
   over which that metric's signal accumulates.

When you re-run this on your own data, expect the picture to look different,
both because the metrics you tracked are different and because individual
patterns vary a lot. The honest interpretation is that this model surfaces
correlations in *your* observed window; it isn't a universal claim about
how sleep affects productivity.

---

# Exercises: Day-of-Week Analysis

Every metric we've looked at so far has been treated as a pure time series:
we've asked whether yesterday, the last 3 days, or the last week predict
today. But daily life has a weekly rhythm. Sleep on Saturday night tends
to look different from sleep on Tuesday night. Stress on Monday morning
tends to look different from stress on Friday afternoon. The `Day` column
in the dataset lets us investigate those patterns directly.

Work through the exercises below using the class `metric_data` (or your
own if you've switched the toggle). Each exercise builds on the previous
one, so work in order.

### Exercise 1: Distribution of a metric by day of the week

Build a boxplot of `Sleep_Hours` faceted or grouped by `Day` (Monday
through Sunday). Use `theme_hpa()` and fill the boxes with `accent_blue`.

Then answer:

  a. Which day has the highest median sleep?
  b. Which day has the widest spread (largest interquartile range)?
  c. Are any days visibly skewed (median far from the mean)?

*Hint:* `metric_data$Day` is a character column. To get the days in
calendar order on the x axis rather than alphabetical, convert it to a
factor with levels in the order you want:
`factor(metric_data$Day, levels = c("Monday", "Tuesday", ..., "Sunday"))`.

### Exercise 2: Comparing weekday and weekend

Create a new column `is_weekend` that is `TRUE` when `Day` is Saturday or
Sunday and `FALSE` otherwise. Then produce a side-by-side density plot
(using `geom_density` with `fill` mapped to `is_weekend`) for two metrics
of your choice.

Answer:

  a. Does the weekday/weekend difference look bigger for morning metrics
     (Sleep_Hours, Wake_Up) or evening metrics (Productivity, Energy)?
  b. Run a two-sample t-test (`t.test()`) comparing weekday vs. weekend
     values for one of the metrics. Is the difference statistically
     significant? What's the mean difference in practical terms?
  c. Why might "statistically significant" and "practically important"
     give different answers on this kind of data?

### Exercise 3: Day-of-week effects in the residuals

The predicted-vs-actual plot earlier showed where the model struggles.
If day-of-week patterns matter and our features don't capture them, the
model's errors (residuals) should depend on the day.

  a. Compute residuals as `residuals <- response - preds`.
  b. Build a data frame joining the residuals back to the `Day` column
     from `metric_data` (careful: you dropped rows where the response
     was missing, so indices need to match up). A safe way is to
     re-filter `metric_data$Day[!is.na(metric_data[[response_metric]])]`.
  c. Plot a boxplot of residuals by Day.
  d. If a particular day has residuals systematically above or below
     zero, the model is biased on that day. Which days show this, if
     any? What feature could you add to capture it?

### Exercise 4: Lagged features with day-of-week interaction

Propose (but you don't need to fully build) a modeling approach that would
let the effect of `mean_val_Sleep_Hours_7` depend on whether the prediction
day is a weekday or a weekend.

  a. One approach: add the interaction explicitly, e.g. create a column
     `mean_sleep_weekend = mean_val_Sleep_Hours_7 * is_weekend`.
  b. Another approach: fit separate models for weekdays and weekends and
     compare their SHAP outputs.
  c. What are the trade-offs between these two approaches? Think about
     sample size in particular, because weekends give you roughly 2/7 of
     the rows.

### Exercise 5 (stretch): Writing up a finding

Pick one day-of-week pattern you discovered in Exercises 1-3 and write
a two-sentence summary in the FIRE framework (Finding, Implication,
Recommendation, Evidence). Keep it concrete: "I noticed X" isn't a
finding until you can say how big X is and how confident you are in it.

